Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-02-26 paper/manuscript.tex
Add my \sbnote{} macro  On branch sboresch  Changes to be committed: 	modified:   manuscript.tex
Adapted/inserted dual coordinate section from gdrive  Changes to be committed: 	modified:   manuscript.tex 	modified:   sb.bib
NEXT: also mention the other refs Emilio sent in dual coordinate section
THEN: naming disambiguation

@@

When operating with a single set of coordinates for the region undergoing the alchemical transformation, an additional distinction between approaches concerns the atom mapping between end state molecules. %Often this is predetermined by the choice of simulation software.
As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}. The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.

@@

As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.

@@

One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.

Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach. Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both). 
To our knowledge, efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.

>>>>

% aux. comments Stefan Boresch -> Stefan Boresch
\newcommand{\sbnote}[1]{%
  {\bfseries{}[SB: }%
  {\textcolor{blue}{#1}}{\bfseries{}]}
  }

@@

When operating with a single set of coordinates for the region undergoing the alchemical transformation, a key distinction between approaches concerns the atom mapping between end state molecules. %Often this is predetermined by the choice of simulation software.
As first suggested by Pearlman, these have been referred to as \emph{dual topology} versus a \emph{single topology} \cite{Pearlman_1994}. The distinction between single and dual topology can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; in particular, consider a transformation of benzene to benzyl alcohol shown in Fig.~\ref{fig:fig_topology}.
In this case the common substructure between the two molecules is the benzene ring, though in practice substructure may be selected to be larger depending the mapping chosen, as we discuss below.

@@

\sbnote{This probably needs to be rewritten\ldots} As far as we are aware, the distinction between single and dual topology approaches is made primarily for two main reasons. First, this choice affects the alchemical pathway followed, and thus may affect convergence properties---though we are not yet aware of a study of the relative efficiency and merits of these two approaches. Additionally, historically, some simulation packages implemented only one approach and not the other, meaning that the distinction was functional.

\paragraph{Dual coordinate approaches --- dual chemical topologies}

The chimeric constructs needed in single and dual topology approaches in the sense of Pearlman’s definition just described quickly become non-trivial to set up and require care to avoid artifacts \sbnote{Ssee below; where do we place the dummy atom discussion?}. Some of these difficulties might be avoided by having the initial and final state present as separate molecules with separate coordinate sets. To the best of our knowledge, such an approach was first described by Axelsen and Li \cite{Axelsen_1998}. Both initial and final state are present simultaneously; the two copies never see each other and interactions with the environment are turned on/off as a function of $\lambda$. Harmonic restraints between corresponding atoms in the initial and final state copy prevent the two molecules (sets of atoms from drifting apart). %Unfortunately, Axelsen and Li referred to their ansatz also as dual-topology, though in practice it is quite different from Pearlman’s meaning of the term.
A recent implementation of this idea is Restraintmaker \cite{Ries_2022}, a tool to automatically assign the distance restraints needed to keep the two molecules / regions (one for each of the physical endstates) on top of each other. Somewhat unfortunately, Axelsen and Li, as well as Ries et al refer to their approach as dual topology. Ries et al suggested to use the term hybrid topology for (single coordinate) dual topology in the sense of Pearlman. We will attempt a disambiguation of the various nomenclatures below.

Another variant of a dual coordinate approach is implemented in NAMD \cite{jiang2019computing}. Whereas Ries et al.\  use distance restraints to keep two molecules on top of each other, corresponding atoms in NAMD are held on top of each other by constraints; in practice, this means that these coordinates are forced to be identical. AMBER's free energy implementation recently switched to a quite similar approach \cite{Lee_2023}.
%The authors refer to this approach as hybrid single-dual topology. While it technically uses two separate sets of coordinates, the constraint that corresponding atoms have identical coordinates makes it effectively a single coordinate method; hence, I consider NAMD’s approach to be a variant of Pearlman’s (single coordinate) dual topology.

Dual coordinate approaches with chemical topologies encompass a third approach, where essentially two absolute free energy calculations are carried out in opposite directions at the same time.
The so-called ``separated topologies'' (SepTop) method \cite{rocklin2013separated,Baumann_2023} and the ``alchemical transfer method'' (ATM) \cite{Azimi_2022} belong into this category. In SepTop, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to (as opposed to Restraintmaker, where the molecules are restrained to each other. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space.

%Other methodologies also use both end states simultaneously; examples are the Separated Topologies (SepTop) approach [Baumann2023] and the ATM method [Azimi2022]. In the former, the two copies are held on top of each other via restraints with respect to the protein/receptor they bind to. Ries et al. refer to such an approach as “separated dual topology”, whereas they label their approach as “linked dual topology”. In ATM, a restraint is used to keep the bound ligand 1 and the free ligand 2 always some minimum distance apart; during exchange moves, suitable rigid-body rotations/translations are applied to move the free ligand into a position/orientation as similar as possible to the one of the bound ligand. Both copies are present simultaneously, but they do not occupy the same space.

%One final approach, so far in its infancy, has been called ``separated topologies'' and essentially consist of two absolute free energy calculations in opposite directions at the same time, turning one molecule's interactions with the environment off, while turning the other molecule's interaction on~\cite{jiang2019computing, rocklin2013separated}.

\paragraph{Summary, comparison of single and dual coordinate methods, orthogonal classifications:}

@@

\sbnote{I think this should get its own paragraph/subsection; to be expanded!} Software packages vary in their use of single or dual topology approaches; for example, AMBER TI uses a dual topology approach, while BioSimSpace uses a single topology approach. Please make sure to check what approach is used with your software package of choice, or whether it supports your choice of approach (GROMACS and GROMOS, for example, support both). 
To our knowledge, efficiency differences have not been thoroughly explored, though conventional wisdom suggests that fewer dummy atoms are better, as introducing or removing atomic sites is usually more difficult, requiring more intermediate steps ~\cite{liu2013lead, mobley2012perspective}.
